Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-08-08 analysis/paper/paper.Rmd
update code and summary of disparity test

To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is openly available online at https://doi.org/10.17605/OSF.IO/ABVGF. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

@@

For the degree of shape standardization across periods, our CV tests on PC1 indicate that there are significant differences between Chinese contact and either pre-European or post-European periods. A test of morphological disparity also demonstrates a similar result as well (See Online Supplementary Materials), suggesting that pottery shape became more standardized after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period. Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size. This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even at small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].

Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization reflected by stylish and functional attributes [@Costin2001; @Costin1995]. In other words, mechanical standardization is defined by potter's skills and experiences that increased morphological uniformity normally relates to fewer producers. Population could be an indicator to estimate the potential number of producers. According to the historical records [@Nakamura1938, pp. 12], the population in Kiwulan reached a higher number in the 17th century as the European presence and decreased in the Chinese period due to movement of indigenous people to the south [@Chen2007]. The population dynamics in historical documents corresponds to an increase in ceramic abundance in the European presence and a decrease in the Chinese period. Thus, we model cv values as a function of the mass of ceramics using Poisson GLM with a link function. The result suggests that ceramic abundance strongly predicts the cv values (β = $`r beta_print`$, p = $`r pval_print`$). This indicates that standardized ceramics in the Chinese period may be influenced by a small population leading to fewer producers if ceramic abundance can reflect the population at Kiwulan.

>>>>

To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is openly available online at https://doi.org/10.17605/OSF.IO/ABVGF [@WangandMarwick2020]. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the visualizations and tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

@@

```{r morphological-disparity}
require(dispRity)
phases_list <-
  list("Pre-European" = which(op.p$phase == "Pre-European"),
       "Post-European" = which(op.p$phase == "Post-European"),
       "Chinese" = which(op.p$phase == "Chinese"))

disparity_data <-
  dispRity.per.group(op.p$x,
                     group = phases_list,
                     metric = c(sum, variances))

disparity <-
  test.dispRity(disparity_data,
                test = wilcox.test,
                correction = "bonferroni",
                details = TRUE)

pre_chi <- signif(disparity[[2]][[1]]$p.value, 4)
post_chi <- signif(disparity[[3]][[1]]$p.value, 4)
```

@@

For the degree of shape standardization, our CV tests on PC1 indicate a significant difference between the Chinese period and either pre-European or post-European periods. A test of morphological disparity also demonstrates similar results between Chinese contact and either pre-European (p = $`pre_chi`$) or post-European periods (p= $`post_chi`$) [@WangandMarwick2020], suggesting that a more standardized pottery shape after contact with the Chinese. In addition, we found a more homogeneous shape accompanied by standardized size, represented by the body diameter of vessels, in the Chinese period. Humans tend to make mistakes in hand-crafting when the size of an object increases, leading to higher variations in larger artifacts [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we use body diameter as our proxy variable of size. This might hint an intentional behavior to achieve a homogeneous form of vessels, but we still need more metric variables to assess the variation in size. Mineral composition shows that the clay pastes are similar throughout three phases, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. We note that a small sample size in the Chinese period may lead to a more standardized shape. However, this effect can be reduced using CV that scales variation to magnitude, allowing reliable comparison across uneven sample numbers, even for small sample size [@Eerkens2001]. Moreover, the MSLR test for equality of CVs enables a robust test between different sample numbers [@Krishnamoorthy2014].

Whether standardization suggests an emergence of specialized production depends on the number of producers, which distinguishes mechanical standardization from intentional standardization reflected by stylish and functional attributes [@Costin2001; @Costin1995]. In other words, mechanical standardization is defined by potter's skills and experiences that increased morphological uniformity normally relates to fewer producers. Population could be an indicator to estimate the potential number of producers. According to the historical records [@Nakamura1938, pp. 12], the population in Kiwulan reached a higher number in the 17th century as the European presence and declined in the Chinese period due to movement of indigenous people to the south [@Chen2007]. The changes in population size described in documents corresponds to the observation of ceramic abundance at Kiwulan. Thus, we model cv values as a function of the mass of ceramics using Poisson GLM with a link function. The model suggests that ceramic abundance strongly predicts the cv values (β = $`r beta_print`$, p = $`r pval_print`$). This indicates that a more standardized shape in the Chinese period may be influenced by a small population if ceramic abundance can reflect the population size.
